---
title: "R fundamentals workbook"
author: "Horacio Gomez-Acevedo"
format: html
bibliography: mainref.bib
editor: visual
---

## An introduction to R

For this introduction we will follow closely the book from @maindonald with some technical details from @chambers

### The console

The command line prompt (\>) is waiting to receive commands or expressions for R to *evaluate* immediately after the keyboard key *Enter* is pressed. The response of R is to evaluate and prints out the result of the evaluation, but in some cases opens a new window with the printout of this process. For instance,

```{r}
#| echo: true
1+3 # a simple sum
# an actual function sum(1,3) is evaluated by R
```

The printout of \[1\] represents that the output consists of only one element (technically of one *object*). Anything after the character \# is considered to be a comment and will not be evaluated nor ignored.

In terms of the interaction with this *notebook*, we can choose to suppress the print out of the following chunk, and only the output is produced.

```{r}
#| echo: false
1 +3
# In this case nothing is print out from this chunk except the result
```

### Simple data entry

Let's consider a modified table from @andersonmay (the last two columns are not part of the original data and the last column contains made up numbers)

```{r}
#| echo: false
place<-c("Cirencester, Enngland", "England and Wales","Kansas, USA", "Ontario, Canada","Willesden, England", "Ghana", "Eastern Nigeria")
timeperiod<-c("1947-1950", "1950-1968","1918-1921","1912-1913","1912-1913","1960-1968","1960-1968")
r0<-c("11-14","16-18","5-6","11-12","11-12","14-15","16-17")
faker0<-c(12.2,17.9,5.5,11.5,11.2,14.8,16.1)
duration<-c(3,18,3,1,1,8,8)
measles.df<-cbind(place,timeperiod,r0,duration,faker0)
colnames(measles.df)<-c("Location","Time period","R0","Duration (years)","Fake R0")
knitr::kable(measles.df,caption = "Historical Estimations of R0 for measles")
```

We can manually insert the data as follows

```{r}
faker0<-c(12.2,17.9,5.5,11.5,11.2,14.8,16.1)
duration<-c(3,18,3,1,1,8,8)
plot(faker0 ~duration, pch=16)
```

Some explanations are in order:

-   The symbol "\<-" is an assignment operator. It means that whatever is defined in the right hand side is assigned to the left hand side. It seems pedantic to use "\<-" instead of "=", but object manipulation in R is different than other computer programs. My suggestion is to use "\<-".

-   The objects *duration* and *faker0* represent "vectors" that are concatenated.

-   The function *plot* will do exactly that we want.

-   The notation inside the function *plot* (i.e., "faker0 \~ duration") is another special type of object referred to as *formula*.

-   The option *pch* referst to the plot character that in this case is just a simple black dot.

This is not our finest representation of a graph, but it shows R functionality.

### The concept of a data frame

Generally speaking, we assume that we have things that can be observed or measured and we call those *variables* (or *covariates*), and also we have a meaningful way to repeat these observations of those variables and report them. In computer sciences, the process of placing data in a consistent and meaningful way is delegated to a *data frame*.

Thus, a data frame provides a framework in which collected data from spreadsheets or databases can be placed in a consistent manner to gain data insights using statistical methodologies.

In R, a data frame object (called nor surprisingly *data.frame*) is a named list, with elements of the list representing variables. Thus, each element of the list should represent the same set of observations.

> A data frame object in R can be tought as a two-way array with columns representing variables.

## A quick overview of R (interactively)

This section we will follow the commands presented in @maindonaldbooklet booklet starting on Chapter 2 section 2. You can check the previous section, the book is very well written.

First, we start defining character vectors based on the table 2.1

```{r}
volume <- c(351 , 955 , 662 , 1203 , 557, 460)
weight <- c(250 , 840 , 550 , 1360 , 640, 420)
```

Next, we put a character vector

```{r}
description <- c("Aird 's Guide to Sydney ",
"Moon 's Australia handbook ",
" Explore Australia Road Atlas",
" Australian Motoring Guide",
" Penguin Touring Atlas", " Canberra - The Guide")
```

> Note. By design, R is optimized to work with vectors. The term *vectorizing* contrasts with the use of loops commonly used in other computer programs (e.g., python).

```{r}
ls(pattern = "on$")
```

We will access specific elements of a vector by specifying their location

```{r}
volume[2]
```

> Note. Unlike other programs, vector positions in R start from 1.

This is a rather odd-looking expression

```{r}
weight/volume
```

What R is doing behind scenes is applying a *vectorized* division. The book has even the application of a function to a vector. Once again, it applies the function to each element of the vector!

```{r}
round(x=weight/volume,digits=3)
```

We have done a simple plot before, let's explore a bit more.

```{r}
plot(volume , weight , pch =16, cex =1.5)
```

Another improvement

```{r}
plot( weight ~ volume , pch =16 , cex =1.5 ,
xlab=" Volume ( cubic mm)", ylab=" Weight (g)")
```

What about the following change

```{r}
plot( weight ~ volume , pch =16 , cex =1.5 ,
xlab=" Volume ( cubic mm)", ylab=" Weight (g)")
identify ( weight ~ volume , labels = description )
```

Let's introduce now *ggplot2*, but before doing that we must keep in mind what type of arguments are required <https://ggplot2.tidyverse.org/reference/ggplot.html>

The first thing we noticed is that *data* should be passed as a *data.frame* object.

```{r}
simple.df<-cbind(volume,weight)
colnames(simple.df)
is.data.frame(simple.df)
```

It looks strange that something that *looks* like a data frame is not the right type of object. We will fix this by forcing (*coercing*) that data structure into a data frame

```{r}
simple.df<-as.data.frame(simple.df)
```

The next chunk will make the plot

```{r}
library(ggplot2)
ggplot(data=simple.df,mapping = aes(x=volume,y=weight))+geom_point()
```

Alternatively, we can put it as **tibbles** using *tidyr* package. You can think of a tibble as a generalization of a data frame in R.

```{r}
library(tidyr)
simple.tibble<-tibble(volume,weight)
ggplot(data=simple.tibble,mapping = aes(x=volume,y=weight))+geom_point()
```

## More on data frames

Let's explore more on data frames using the book example

```{r}
travelbooks <- data.frame (
thickness = c(1.3 , 3.9 , 1.2 , 2, 0.6 , 1.5),
width = c(11.3 , 13.1 , 20, 21.1 , 25.8 , 13.1),
height = c(23.9 , 18.7 , 27.6 , 28.5 , 36, 23.4),
weight = weight , # Use values entered earlier
volume = volume , # Use values entered earlier
type = c(" Guide ", " Guide ", " Roadmaps ", " Roadmaps ",
" Roadmaps ", "Guide"),
row.names = description
)
## Remove objects that are not now needed.
rm(volume , weight , description )
```

As a list of vectors, we will have some flexibility for accessing individual *cells*, rows, columns. However, it takes some time to get familiar with conventions for *slicing* the data according to our needs.

-   Accessing specific row

```{r}
travelbooks[5,]
```

-   Accessing specific column (by name)

```{r}
travelbooks[,"volume"]
```

The conventional to access specific column has two different looking expressions

```{r}
travelbooks[["width"]]
```

And,

```{r}
travelbooks$width
```

There are more options, such a selecting specific rows based on a condition.

```{r}
travelbooks[travelbooks$height>30,]
```

## Input data from a file (part 1)

We will use the package [DAAG](https://cran.r-project.org/web/packages/DAAG/index.html)

```{r}
library(DAAG)
datafile("travelbooks")
```

If you check the file is similar to our previous data frame, and it writes out in our project directory.

```{r}
my.travel.books<-read.table(file="travelbooks.txt",header = TRUE,row.names = 1)
```

# In class exercise

I have placed a small data set called Duke Students GPA in the class folder from [Kaggle](https://www.kaggle.com/datasets/joebeachcapital/duke-students-gpa)
