---
title: "Data Cleaning"
author: "Horacio Gomez-Acevedo"
format: html
bibliography: ../mainref.bib
editor: visual
---

## Tibbles

We will follow the online version of the book R for Data Science <https://r4ds.had.co.nz/tibbles.html>

Tibbles are a new object that act as an *advanced* version of a data frame. The expansion of a data frame traditional structure allows us to clean data faster and more efficiently.

```{r}
library(tidyverse)
```

> If you see in your system a message as above, there is no need to panic. It means that are *conflicts* because there are functions with the same name coming from different packages.

We will begin with the traditional data frame *iris*. Here we just print the data frame and check that it is indeed a *data.frame* object.

```{r}
head(iris)
print(class(iris))
```

Now, we convert this data frame to a *tibble*

```{r}
t.iris<-as_tibble(iris)
t.iris
```

### Building tibbles

```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

What are the differences?

- a tibble is not adding a row number
- a tibble does not change the name of the variables
- a tibble does not convert strings into factors

There is a nice function to *transpose* a tibble called *tribble*. It is useful for getting data entry within your code. Column headings start with a *formula* (meaning that they should include "\~" symbol)

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

### Data extraction from tibbles

Normal operations that we have used already in data frames

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)
df
# Extract by name
df$x
#or
df[["x"]]

# Extract by position
df[[1]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
# Contrast this with 
df[,1]
```

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
annoying
```

Some more examples

```{r}
a <- 1:5
tibble(a, a ** 2)
```

The printout shows something interesting, the name of the second column is assigned as \* a\^2 \*. But this is a valid name for tibbles!

We can play a bit more

```{r}
tibble(a, b = a * 2, c = 1,`a+b`=a+b)

```

### Adding rows and columns

The basic function works as follows

```{r}
df <- tibble(x = 1:3, y = 3:1)
df<-add_row(df,x=4,y=10)
```

However, it is more efficient using the so-called **pipelines** that are used widely in tidyverse

```{r}
df %>% add_row(x = 5, y = 10)
```

There is more functionality that we can use with \*add_row\*\*

```{r}
# You can specify where to add the new rows
df<-df %>% add_row(x = 6, y = 0, .before = 2)
```

Something similar is for adding columns

```{r}

df<-df %>% add_column(z= 1:length(df$x))
```

## Tidyr

There is a very nice overview of what is *tidy data*

1.  Each variable forms a column
2.  Each observation forms a row
3.  Each type of observational unit forms a table

*Tidyr* is part of *tidyverse* <https://tidyr.tidyverse.org/>. Run in your console

```{r}
vignette("tidy-data")
```

We will use parts of the data that appears on the paper @tidydatap

```{r}
billboard
```

Then, if we use *pivot_longer* we can put that data frame in a better shape

```{r}
billboard2 <- billboard %>% 
  pivot_longer(
    wk1:wk76, 
    names_to = "week", 
    values_to = "rank", 
    values_drop_na = TRUE
  )
billboard2
```

Moreover, we can remove the *wk* from each week for analysis and adjusting for the exact week!

```{r}
billboard3 <- billboard2 %>%
  mutate(
    week = as.integer(gsub("wk", "", week)),
    date = as.Date(date.entered) + 7 * (week - 1),
    date.entered = NULL
  )
billboard3
```

Sorting by artist, track and week

```{r}
billboard3 %>% arrange(artist, track, week)
```

### Cheatsheet

This is very useful <https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf>
