---
title: "Statistical Review"
author: "Horacio Gomez-Acevedo, PhD"
format: html
bibliography: mainref.bib
editor: visual
---

This lecture is based on @maindonald book chapter 3 and @harrell chapter 2

## General Overview


In general, an statistical model relies on probabilistic forms of description of events that have an application in science. They normally consist of a deterministic component and a random component that attempts to capture experimental variability.

There are a number of models that can be used and it is critical to carry on a formal and scientific analysis. Models' choice may depend on experience with the data itself, subject area knowledge, and sometimes motivated by the EDA.

A simple example is given with the DAAG package

```{r}
library(DAAG)
attach(roller)
```

In this data set the variable *depression* for different *weights*  of lawn roller are given. Assuming a line through the origin the model can be represented as

$$
\text{depression}=\beta* \text{weight}+ \text{noise}
$$
where $\beta$ needs to be estimated, and the noise is different for each different part of the lawn.

Mathematically we represent here as 

$$Y= \mu + \varepsilon$$

The model prediction $\mu$ is the signal component of the model, whereas $\varepsilon$ is the noise or statistical error component. Sometimes, we use the idea of the *expected value* and it is normal to define $\mu=E(Y|X)$. 

## Models

The above model treats the pattern of change of depression with roller weight as a *deterministic* or *fixed* effects. The measured values of depression incorporate, in addition, a *random* effect that reflects variation from one part of the lawn to another. 



## Fitting models

In R, we use a *formula* expression to suggest the relationship

```{r}
roller.lm <- lm (depression ~ weight, data=roller)
plot(depression~ weight,data=roller)
abline(roller.lm)
```

Note that the model object *roller.lm* is a list

```{r}
names(roller.lm)
```

Then, you can extract extra information

```{r}
roller.lm$coefficients
```

Then, if you want to extract the estimated value of $\beta$ you can use the command $\beta$=`r roller.lm$coefficients['weight'] `

## Regression Models

We will generalize the above ideas to accommodate future models.

Simply stated, a regression model is defined in terms of a connection between the predictors $X$ and the response $Y$. If $C(Y|X)$ denote a property (or function) of the distribution of $Y$ given $X$ (our *covariates*).

In the above example 

$$
C(Y|\text{weight})= \beta *\text{weight} + \beta_0 
$$

Moreover, this ordinary linear regression model the expected or average of $Y$ given *weight* or 

$$
C(Y|\text{weight})=E(Y|\text{weight})
$$
and it is assumed that $Y$ has a normal distribution with mean $\beta$ and constant variance $\sigma^2$. 

A more general linear regression model can be given by

$$
C(Y|X)= g(X\beta)
$$
where $g$ is a function (oftentimes nonlinear) and $X \beta$ represents a weighted sum of all the $X$s.

For instance for regression models we will see that 

$$
C(Y|X)= \text{Prob}(Y=1|X)= \frac{1}{1+ \exp(-X\beta)}
$$

Sometimes, it is better to isolate $X\beta$ through what is commonly called *link function*. In the previous example if we define $h(u)=\log (u/(1-u))$

$$
h(C(Y|X))=h(\text{Prob}(Y=1|X))= \log(\exp(X\beta))=X\beta
$$

Thus, if we denote $C'(Y|X)=h(C(Y|X))$ a **general linear model** can be written as

$$
C'(Y|X)=X\beta
$$

The left hand side of the previous equation means that we have done some transformation with $Y$, whereas the right hand side is the *linear regression* part. 

## Interpreting Model Parameters

If we consider the model 

$$
C'(Y|X)= X\beta= \beta_0+ \beta_1 X_1 + \cdots + \beta_p X_p
$$

the regression parameter $\beta_j$ is interpreted as the change in the property $C'$ of $Y$  per unit change in the descriptor variable $X_j$ when all other descriptors remain constant

$$
\beta_j= C'(Y|X_1,X_2,\ldots,X_j+1,\ldots,X_p)
-C'(Y|X_1,X_2,\ldots,X_j,\ldots,X_p)
$$

+ In ordinary linear regression, $\beta_j$ is the change in expected value of $Y$ per unit change in $X_j$.
+ In logistic regression, $\beta_j$ is the change in log odds that $Y=1$ per unit change in $X_j$.

## References