---
title: "Introduction to ggplot2"
author: "Horacio Gomez-Acevedo"
institute: "University of Arkansas for Medical Sciences"
format: html
bibliography: ../mainref.bib
editor: visual
---

## What is ggplot2?

[ggplot2](https://ggplot2.tidyverse.org/) is an R package for producing statistical and data graphics. *ggplot2* has its own grammar based on the **Grammar of Graphics** and is composed of independent components that can interact by themselves and produce very impressive plots.

The installation is typical in R.

```{r install, eval=FALSE}
install.packages("tidyverse")
```

## Grammar of Graphics

In 2005, Leland Wilkinson published a book entitled *The Grammar of Graphics* @grammargraph in which he aimed to answer the question: **what is a statistical graphic?**

Build upon the layered grammar of graphics, Hadley Wickham ports these concepts into R and published @ggplot2book in which provides the following definition

> A statistical graphic is a mapping from data to some aesthetic attributes of geometric objects.

-   The plot may also contain statistical transformations of the data
-   Faceting can be used to focus in different subsets of the data
-   Thus, a graph consists of a combination of the previous components.

## Code

This is a typical example of ggplot2 in action

```{r,fig.width=10,fig.height=5}
library(ggplot2)
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```

## Building of a plot

To create the *plot* object, we use the function **ggplot()**. It has two arguments:

-   data
-   aesthetic **mapping**

```{r}
library(ggplot2)
#p<-ggplot(diamonds, aes(carat, price, colour=cut))
```

Why we don't see anything? It is because it cannot be displayed unless it has a **layer**

A layer combines data, aesthetic mapping, a geom (geometric object), a stat (statistical transformation), and a position adjustment.

> Despite that @ggplot2book labels some of these parts as optional, recent versions of ggplot2 make it mandatory to have at least the *geom*, *stat* and *position* specified.

```{r}
p<-ggplot(diamonds, aes(x=carat))
p<-p+layer(geom="bar",
           stat="bin",
           position = position_identity()
)
p
```

Behind the scenes, the functions *ggplot()* and *qplot()* creates a plot object and add layers that are defined. Thus, it is normal in ggplot2 to see the '+' as a way to add a specific layer.

A much simpler way to create layers is by using a geom\_ function.

```{r}
q<-ggplot(diamonds, aes(x=carat))
q+geom_histogram()
```

## Ggplot main parameters

-   Data. The data option must be filled with a *data frame*.
-   Aesthetic mappings. The *aes* function describes the way that variables in the data are mapped to graphical objects in the plot.

For instance

```{r, eval=FALSE}
aes(x=heigh, y=weight, colour=age)
```

The *colour=age* refers to a factor that will contain such information.

-   Geometric objects or *geoms* perform the actual rendering of the layer and control the type of plot.

## Ggplot optional parameters

-   Statistical transformation or *stat* normally modify the data. Typically it summarizes the data.\
    For example the *stat_bin* statistic used for histograms produces the following variables
    -   count, the number of observation in each bin
    -   density, the density of observations in each bin
    -   x, the center of the bin
-   Position. This parameter allow us to perform minor tweaks to the position elements within a layer.

## Examples

Let's consider the data coming out from the *Duke Students GPA* data set named *gpa.csv*

```{r}
h.data<-read.csv(file = "../Datasets/gpa.csv",header = T,sep = ",")
```

The variables in this data set are:

-   gpa
-   studyweek. The number of study hours per week
-   sleepnight. The number of sleep hours per night
-   out. The nights going out per week.
-   gender

### Basic EDA

Let's get an idea of the distribution of the variable gpa.

```{r}
pgpa<- ggplot(data=h.data,aes(x=gpa))
pgpa + geom_density(fill="pink", alpha=(alpha=0.3))
```

Let's investigate if the distribution changes based on gender.

```{r}
f.students<-h.data[h.data$gender=="female",]
m.students<-h.data[h.data$gender!="female",]
```

We already see that the sample sizes are different: males `r dim(m.students)[1]` and females `r dim(f.students)[1]`

```{r}
pgpa.all<- ggplot(data=h.data,aes(x=gpa,fill=gender,colour=gender))
pgpa.all + geom_density(alpha=0.3)
```

Let's check for possible outliers

```{r}
pbp<-ggplot(data=h.data,aes(gender,gpa))
pbp+geom_boxplot(fill="purple",colour="#c51b8a",alpha=0.2,outlier.color = "blue",outlier.shape = 1,outlier.alpha = 1)
```

## Geoms

Geometric objects, or **geom** for short, perform the actual rendering of the layer, control the type of plot that is requested.

Each *geom* has a set of aesthetics that it understands, and a set that are required for drawing.

Every *geom* has a default statistic, and every statistic a default geom.

## Stats

A statistical transformation, or **stat** for short, transform the data, typically summarizing it in some manner.

A *stat* takes a data set as input and returns a data set as output, and so a stat can add new variables to the original data set.

```{r}
library(ggplot2)
d<-ggplot(data=diamonds,aes(carat))+xlim(0,3)
d+stat_bin(aes(ymax=after_stat(count)),bindwidth=0.1,geom="area")
```

```{r}
d+stat_bin(aes(size=after_stat(density)), geom="point",position="identity" )
```

## Scales

Often times we need to make transformations in our data. Here are two mechanisms by which we can make that happen.

```{r}
p<-ggplot(data=diamonds,aes(log10(carat),log10(price)))
p+geom_point()
```

or

```{r}
p<-ggplot(data=diamonds,aes(carat,price))
p+geom_point()+scale_x_log10()+scale_y_log10()
```

## Color

There is a more thorough description of color description in this website <https://www.handprint.com/HP/WCL/color7.html> .

We will use color gradients and those come in three flavors:

-   *scale_color_gradient()* and *scale_fill_gradient()* give a two-color gradient low to high.

-   *scale_color_gradient2()* is a three color gradient, low medium and high.

-   *scale_color_gradientn()* is a custom color that requires a vector of colors as argument.

We will create a 2d density estimate of the *faithful* data set.

```{r}
# we will create the data frame that is required
fd2<-with(faithful,MASS::kde2d(eruptions,waiting, h=c(1,10),n=50))
df<-with(fd2,cbind(expand.grid(x,y),as.vector(z)))
names(df)<-c("eruptions","waiting","density")
erupt<-ggplot(df,aes(waiting, eruptions, fill=density))+geom_tile()+scale_x_continuous(expand=c(0,0))+scale_y_continuous(expand=c(0,0))
erupt+scale_fill_gradient(limits=c(0,0.04))
```

But we can use more

```{r}
erupt+scale_fill_gradient2(limits=c(-0.04,0.04),midpoint=mean(df$density))
```

### Colorbrewer

We can also use colorbrewer colors <https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3> and the library needs to be installed.

```{r}
library(RColorBrewer)
point<-ggplot(data = msleep,aes(brainwt,bodywt,color=vore))+geom_point()+scale_x_log10()+scale_y_log10()
point+scale_color_brewer(palette = "Set1")
```

```{r}
point+scale_color_brewer(palette = "Pastel1")
```

## Faceting

Faceting generates small multiples each showing a different subset of the data.

> Note. It only works when you are using the same data set.

The default is that you get a single plot

```{r}
mpg2<-subset(mpg,cyl !=5 & drv %in% c("4","f"))
qplot(cty,hwy, data=mpg2)+facet_grid(. ~ .)
```

A single row with multiple columns based on the levels of *cyl*.

```{r}
qplot(cty,hwy,data=mpg2)+facet_grid(.~cyl)
```

If you want a single column with multiple rows

```{r}
qplot(cty,data=mpg2,geom="histogram",bindwidth=2)+facet_grid(cyl~.)
```

For multiple rows and columns

```{r}
qplot(cty,hwy,data=mpg2)+facet_grid(drv ~ cyl)
```

## References

::: {#refs}
:::
